Faithful short-read mapping with Sesame

### 2019-09-20 manuscript.tex
Add estimation strategy and start super reads

From this we draw the important conclusion that the performance of skip
seeds degrades much less than MEM seeds when the target contains
paralogues. Skip seeds are thus better suited for sensitive mapping in
repeated sequences. Figure~\ref{fig:mortal_kombat} also shows that the
performance of skip seeds on this criterion is simply better than MEM
seeds. However one has to bear in mind that we compare MEM seeds of size
19 with skip seeds of size 16: the gain in sensitivity comes at the cost
of a substantially larger candidate set. In practice the candidate set
must be further filtered, so the final performance cannot be as high as
suggested by Figure~\ref{fig:mortal_kombat}.

Another interesting observation is that for MEM seeds, the off-target rate
is above $10^{-3}$ for reads of 50 nucleotides or less, meaning that
beating quality 30 is not straightforward in these conditions. It is
important to remember that mapping cannot be wrong if the target has no
paralogue (because it is trivial to recognize that false positives have no
homology with the read). Therefore, the main strategy to reach high
mapping confidence with MEM seeds is not to reduce the minimum seed size,
but to ascertain that the target is a unique sequence. We will show how to
do achieve this below.

\section{Estimating parameters $N$ and $\mu$}

A key component of...

>>>>

From this we draw the important conclusion that the performance of skip
seeds degrades much less than MEM seeds when the target contains
paralogues. Skip seeds are thus better suited for sensitive mapping in
repeated sequences. Incidentally, this Figure~\ref{fig:mortal_kombat}
shows that it is suboptimal for a MEM-based mapper to all the candidate
loci when there is more than a handful of them: if the target has more
than 20 paralogues, the mapping quality will remain low even when the
output is correct. A better strategy is to either bail out or switch to a
more sensitive seeding method (BWA uses a re-seeding policy).

A related observation is that for MEM seeds, the off-target rate is above
$10^{-3}$ for reads of 50 nucleotides or less, meaning that beating
quality 30 is not straightforward in these conditions. It is important to
remember that mapping cannot be wrong if the target has no paralogue
(because it is trivial to recognize that false positives have no homology
with the read). Therefore, the main strategy to reach high mapping
confidence with MEM seeds is not to reduce the minimum seed size, but to
ascertain that the target is a unique sequence. We will show how to do
achieve this below.

Figure~\ref{fig:mortal_kombat} suggests that the performance of skip seeds
is better than that of MEM seeds. However one has to bear in mind that we
compare MEM seeds of size 19 with skip seeds of size 16: the gain in
sensitivity comes at the cost of a larger candidate set. In practice such
a large candidate set must be further filtered (Bowtie2 does this with a
priority policy), so the final performance of skip seeds cannot be as high
as suggested in the figure because of speed constraints.

\section{Estimating $N$ and $\mu$}

Estimating the two read-dependent parameters $N$ and $\mu$ is essential to
compute seeding probabilities. Recall that $N$ is the number paralogues of
the target ($N=0$ means that it is unique) and that $\mu$ is the sequence
divergence (the average fraction of mismatches between a paralogue and the
target).

The key observation is that the seeding algorithm can also reveal the
paralogue structure of the target. Queries in the FM index are carried out
using an algorithm known as the backward search. Extending the query one
nucleotide at a time, the backward search returns the number of matching
sequences in the genome. The decaying number of hits depends on how many
paralogues of the target are present in the genome, and how similar they
are with each other. It is thus possible to use the series of the backward
search to estimate $N$ and $\mu$.

Writing the likelihood of the process is straightforward if we assume that
nucleotides are used uniformly at random in the genome. The difficulty is
that the first iterations of the backward search are uninformative. For
instance, a query of length 10 is expected to have $\geq 3000$ hits in
the genome, so the potential paralogues of the target are hidden among so
many random hits. In practice, we ignore the first iterations until the
number of random hits is lower than 1. The information is labelled as
missing and the likelihood is updated accordingly. This gives a function
that can now by maximized with respect to $N$ and $\mu$.

Initial simulations revealed that the maximum likelihood estimates of $N$
and $\mu$ have a large variance. To mitigate this effect, we opted for a
grid search where we test only plausible values of $\mu$ and compute the
associated $N$ (see Methods). This gave a fast algorithm forced to produce
reasonable estimates.

We initially tested this method on the read, gathering the series of decay
during the seeding process. However, we quickly realized that this method
routinely missed paralogues because of sequencing errors. Using mapping as
an error correction mechanism, we tested a prototype estimating $N$ and
$\mu$ from the sequence of the best hit and obtained significantly better
results, so we settle for this option, even though it is slower. It is of
course possible that the best hit is not the target, but if it is a
paralogue, the estimated values of $N$ and $\mu$ should be similar.

The estimation strategy is depicted in Figure~\ref{fig:strategy}. The
putative target is segmented in windows of 30 nucleotides overlapping by
20 nucleotides. The grid search is run a every window in both the foward
and reverse orientation, ignoring the first 20 iterations of the backward
search. The windowing approach accomodates the fact that only some
regions of the target may be repeated (e.g., if the read straddles a
transposon). Finally, we take the median values of $N$ and $\mu$ as
representative for the whole sequence.

\section{Super reads}

While testing the estimation algorithm on reads of size 50, we observed
that a sub-population of them had mapping error rates below $10^{-5}$.
This is surprising in regard of Figure~\ref{fig:mortal_kombat}a because in
a genome where half of the sequences are duplicated, the error rate should
be $\approx 1/2 10^{-3}$. We reasoned that non-repeated sequences must be
overrepresented in this sub-population of reads.

Those reads are characterized by two facts: $i.$ they have at most one
difference with the best hit and $ii.$ the backward search depicted in
Figure~\ref{fig:strategy} retrived only one sequence on each window.
Individually, these facts are not uncommon in mismapped reads, but
together, they leave little space for error.
